---
title: "Community trajectory analysis"
author: "Miquel De Cáceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Community trajectory analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{vegclust}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction
In this vignette you will learn how to conduct community trajectory analysis (CTA), as implemented in package **vegclust**.

```{r load libraries, echo = T}
library(vegclust)
library(RColorBrewer)
```


## 2. Simple example

In this section we describe how to study the trajectories of two sites that have been surveyed three times each. 

### 2.1 Trajectory data
To specify community dynamics, we need three data items: (a) a set of community states (i.e. coordinates in a space $\Omega$), described using a distance matrix $d$; (b) a vector specifying the site corresponding to each community state; (c) a vector specifying the survey corresponding to each community state. Let's first define the vectors that describe the site and the survey of each community state:
```{r}
#Description of sites and surveys
sites = c(1,1,1,2,2,2)
surveys=c(1,2,3,1,2,3)
```
We then build a matrix with the coordinates corresponding to the community states of the two sites during each survey. We assume that the community space $\Omega$ has 2 dimensions:
```{r}
#Raw data table
xy<-matrix(0, nrow=6, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4:6,1] <- 0.5
xy[4:6,2] <- xy[1:3,2]
xy[6,1]<-1
cbind(sites,surveys,xy)
```
The matrix of Euclidean distances $d$ between community states in $\Omega$ is then:
```{r} 
#Distance matrix
D = dist(xy)
D
```

### 2.2 Display trajectories

To begin our analysis of the two trajectories we display them in an ordination space, using function `trajectoryPCoA`. Since $\Omega$ has only two dimensions, the Principal Coordinates Analysis (PCoA) on $d$ displays the complete space:

```{r pcoa, fig = TRUE, fig.height=4, fig.width=5, fig.align = "center"}
trajectoryPCoA(D, sites, surveys, traj.colors = c("black","red"), lwd = 2)
```

While trajectory of site '1' (black arrows) is made of two segments of the same length and direction, trajectory of site '2' (red arrows) has a second segment that bends and is longer than that of the segond segment of site '1'. 

### 2.3 Trajectory lengths and angles

We can obtain the length of trajectory segments using function `trajectoryLengths` and the angles between consecutive segments, using function `trajectoryAngles`:

```{r} 
trajectoryLengths(D, sites, surveys)
trajectoryAngles(D, sites, surveys)
```
In this case, the same information could be obtained by inspecting the PCoA representation, but in a case of $\Omega$ having many dimensions, the PCoA representation will correspond to a reduced space and hence, angles and lengths will not correspond exactly to those of functions `trajectoryLengths` and `trajectoryAngles`, which take into account the full space.

### 2.4 Relative positions within trajectories
Community states occupy a position within their trajectory that depends on their position along the total pathway of the trajectory. By adding the length of segments prior to a given state and dividing the sum by the total length of the trajectory we obtain the relative position of the community state. Function `trajectoryProjection` allows obtaining the relative position of each point of a trajectory. To use it for this purpose one should use as parameters the distance matrix between state and the indices that conform the trajectory. For example for the two example trajectories we would have:

```{r} 
trajectoryProjection(D, 1:3, 1:3)
trajectoryProjection(D, 4:6, 4:6)
```  
Because the trajectory of site '2' as a second segment that is longer than the others, the relative position of state '5' is lower than 0.5.

Function `trajectoryProjection` can also be used to project arbitrary community states on a given trajectory. For example we can study the projection of third state of the trajectory of site '1' (i.e. state 3) onto the trajectory of site '2' (i.e. states 4 to 6):
```{r} 
trajectoryProjection(D, 3, 4:6)
```  


### 2.5 Distances between segments and trajectories

To start comparing trajectories between sites, one important step is the calculation of distances between directed segments, which can be obtained by calling function `segmentDistances`:
```{r} 
segmentDistances(D, sites, surveys)$Dseg
```  
Distances between segments are affected by differences in both *position* and *direction*. Hence, among the four segments of this example, the distance is maximum between the first segment of trajectory '1' and the second segment of trajectory '2'.

Distances between segments are internally calculated when comparing whole trajectories using function `trajectoryDistances`. Here we show the dissimilarity between the two trajectories as assessed using either the *Hausdorff* distance (equal to the maximum distance between directed segments) or the *directed segment path distance* (an average of distances between segments): 
```{r} 
trajectoryDistances(D, sites, surveys, distance.type = "Hausdorff")
trajectoryDistances(D, sites, surveys, distance.type = "DSPD")
```  

## 3. Structural dynamics in permanent plots

In this example we analyze the dynamics of 8 permanent forest plots located on slopes of a valley in the New Zealand Alps. Forests plots are almost monospecific (*Fuscospora cliffortioides*) and were affected by different disturbances during the sampling period (1972-2009) which includes 9 surveys. We begin by loading the data.
```{r load avoca, echo=T}
data("avoca")
```

```{r distance, echo=TRUE}
avoca_D_man = vegdiststruct(avoca_strat, method="manhattan", transform = function(x){log(x+1)})
```

### 3.1 Display trajectories in PCoA
```{r avoca pcoa, echo=T, fig=TRUE, fig.height=6, fig.width=6, fig.align = "center"}
trajectoryPCoA(avoca_D_man,  avoca_sites, avoca_surveys,
               traj.colors = brewer.pal(8,"Accent"), 
               axes=c(1,2), length=0.1, lwd=2)
legend("topright", bty="n", legend = 1:8, col = brewer.pal(8,"Accent"), lwd=2)

```

### 3.2 Trajectory lengths and angles

```{r trajectory lengths, echo=T}
trajectoryLengths(avoca_D_man, avoca_sites, avoca_surveys)
```


```{r int1, echo=FALSE}
plotTrajDiamDist<-function(cli = 7) {
l = colnames(avoca_strat[[1]])
ncl = 14
m197072= avoca_strat[avoca_surveys==1][[cli]]["NOTCLI",2:ncl]
m197072[m197072<1] = NA
m1974 = avoca_strat[avoca_surveys==2][[cli]]["NOTCLI",2:ncl]
m1974[m1974<1] = NA
m1978 = avoca_strat[avoca_surveys==3][[cli]]["NOTCLI",2:ncl]
m1978[m1978<1] = NA
m1983 = avoca_strat[avoca_surveys==4][[cli]]["NOTCLI",2:ncl]
m1983[m1983<1] = NA
m1987 = avoca_strat[avoca_surveys==5][[cli]]["NOTCLI",2:ncl]
m1987[m1987<1] = NA
m1993 = avoca_strat[avoca_surveys==6][[cli]]["NOTCLI",2:ncl]
m1993[m1993<1] = NA
m1999 = avoca_strat[avoca_surveys==7][[cli]]["NOTCLI",2:ncl]
m1999[m1999<1] = NA
m2004 = avoca_strat[avoca_surveys==8][[cli]]["NOTCLI",2:ncl]
m2004[m2004<1] = NA
m2009 = avoca_strat[avoca_surveys==9][[cli]]["NOTCLI",2:ncl]
m2009[m2009<1] = NA


plot(m197072, type="l", ylim=c(1,200), log="y",
       xlab="", ylab="Number of individuals (log)", main=paste0("Trajectory ",cli), 
       axes=FALSE, col=gray(0.8), lwd=2)
axis(2, las=2)
axis(1, at=1:(ncl-1), labels=l[2:ncl], las=2)
lines(m1974, col=gray(0.7), lwd=2)
lines(m1978, col=gray(0.6), lwd=2)
lines(m1983, col=gray(0.5), lwd=2)
lines(m1987, col=gray(0.4), lwd=2)
lines(m1993, col=gray(0.3), lwd=2)
lines(m1999, col=gray(0.2), lwd=2)
lines(m2004, col=gray(0.1), lwd=2)
lines(m2009, col=gray(0), lwd=2)
legend("topright", bty="n", lwd=2,col=gray(seq(0.8,0, by=-0.1)), legend=c("1970/72","1974","1978","1983", "1987", "1993","1999","2004","2009"))
}
```


```{r trajectory angles, echo=T}
trajectoryAngles(avoca_D_man, avoca_sites, avoca_surveys)
```


```{r trajectory 3 DBH dist, echo=F, fig.height=4, fig.width=8, fig.align = "center"}
par(mfrow=c(1,2))
trajectoryPCoA(avoca_D_man,  avoca_sites, avoca_surveys,
               selection= 3,
               length=0.1, lwd=2)
plotTrajDiamDist(3)
```

```{r trajectory 4, echo=T, fig.height=4, fig.width=8, fig.align = "center"}
par(mfrow=c(1,2))
trajectoryPCoA(avoca_D_man,  avoca_sites, avoca_surveys,
               selection= 4,
               length=0.1, lwd=2)
plotTrajDiamDist(4)
```



### 3.3 Distances between trajectories

```{r avoca DT, echo=FALSE}
avoca_D_traj_man = trajectoryDistances(avoca_D_man, avoca_sites, distance.type="DSPD", verbose=FALSE)
print(round(avoca_D_traj_man,3))
```

```{r avoca DT PCoA, echo=FALSE, fig = TRUE, fig.height=5, fig.width=6, fig.align="center"}
cmd_D2<-cmdscale(avoca_D_traj_man, add=TRUE, eig=TRUE, k=7)
x<-cmd_D2$points[,1]
y<-cmd_D2$points[,2]
plot(x,y, type="p", asp=1, xlab=paste0("PCoA 1 (", round(100*cmd_D2$eig[1]/sum(cmd_D2$eig)),"%)"), 
     ylab=paste0("PCoA 2 (", round(100*cmd_D2$eig[2]/sum(cmd_D2$eig)),"%)"), col="black",
     bg= brewer.pal(8,"Accent"), pch=21)
text(x,y, labels=1:8, pos=2)
```


