---
title: "Community trajectory analysis"
author: "Miquel De Cáceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Community trajectory analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{vegclust}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction
In this vignette you will learn how to conduct community trajectory analysis (CTA), as implemented in package **vegclust**.

We begin by loading the libraries:

```{r load libraries, echo = T}
library(vegclust)
library(RColorBrewer)
```


## 2. Simple example

In this section we describe how to study the trajectories of two sites that have been surveyed four times each. We use a simple example so that geometric calculations can be followed more easily.

### 2.1 Trajectory data
To specify community dynamics, we need three data items: (a) a set of community states (i.e. coordinates in a space $\Omega$), described using a distance matrix $d$; (b) a vector specifying the site corresponding to each community state; (c) a vector specifying the survey corresponding to each community state. Let's first define the vectors that describe the site and the survey of each community state:
```{r}
#Description of sites and surveys
sites = c(1,1,1,1,2,2,2,2)
surveys=c(1,2,3,4,1,2,3,4)
```
We then build a matrix with the coordinates corresponding to the community states of the two sites during each survey. We assume that the community space $\Omega$ has 2 dimensions:
```{r}
#Raw data table
xy<-matrix(0, nrow=8, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4,2]<-3
xy[5:7,2] <- xy[1:3,2]
xy[8,2]<-2.5
xy[5:6,1] <- 0.5
xy[7,1]<-1.0
xy[8,1]<-2
cbind(sites,surveys,xy)
```
The matrix of Euclidean distances $d$ between community states in $\Omega$ is then:
```{r} 
#Distance matrix
D = dist(xy)
D
```

### 2.2 Display trajectories

To begin our analysis of the two trajectories we display them in an ordination space, using function `trajectoryPCoA`. Since $\Omega$ has only two dimensions, the Principal Coordinates Analysis (PCoA) on $d$ displays the complete space:

```{r pcoa, fig = TRUE, fig.height=4, fig.width=5, fig.align = "center"}
trajectoryPCoA(D, sites, surveys, traj.colors = c("black","red"), lwd = 2)
legend("topright", col=c("black","red"), legend=c("Trajectory 1", "Trajectory 2"), bty="n", lty=1, lwd = 2)
```

While trajectory of site '1' (black arrows) is made of two segments of the same length and direction, trajectory of site '2' (red arrows) has a second segment that bends and is longer than that of the segond segment of site '1'. 

### 2.3 Trajectory lengths and angles

We can obtain the length of trajectory segments using function `trajectoryLengths` and the angles between consecutive segments, using function `trajectoryAngles`:

```{r} 
trajectoryLengths(D, sites, surveys)
trajectoryAngles(D, sites, surveys)
```
where some segments are longer angles are smaller and for trajectory of site '2'. In this case, the same information could be obtained by inspecting the PCoA representation, but in a case of $\Omega$ having many dimensions, the PCoA representation will correspond to a reduced space and hence, angles and lengths will not correspond exactly to those of functions `trajectoryLengths` and `trajectoryAngles`, which take into account the full space.

### 2.4 Relative positions within trajectories
Community states occupy a position within their trajectory that depends on their position along the total pathway of the trajectory. By adding the length of segments prior to a given state and dividing the sum by the total length of the trajectory we obtain the relative position of the community state. Function `trajectoryProjection` allows obtaining the relative position of each point of a trajectory. To use it for this purpose one should use as parameters the distance matrix between state and the indices that conform the trajectory. For example for the two example trajectories we would have:

```{r} 
trajectoryProjection(D, 1:4, 1:4)
trajectoryProjection(D, 5:8, 5:8)
```  
Because the trajectory of site '2' as the second and third segments that are longer than the first, the relative position of state '6' is lower than 1/3 and that of state '7' is lower than 2/3.

Function `trajectoryProjection` can also be used to project arbitrary community states on a given trajectory. For example we can study the projection of third state of the trajectory of site '1' (i.e. state 3) onto the trajectory of site '2' (i.e. states 5 to 8):
```{r} 
trajectoryProjection(D, 3, 5:8)
```  

### 2.5 Trajectory directionality and convergence
One way of determining the degree to which trajectories follow trends (i.e. more or less straight pathways) is to calculate the Mantel correlation between the dissimilarity matrix containing distances between points of the trajectories and distance matrix containing the separation in terms of surveys. Surveys are far appart in time they should correspond to a larger distances than surveys that are close in time. The function that calculates the Mantel correlation statistic is `trajectoryDirectionality`:
```{r} 
trajectoryDirectionality(D, sites, surveys)
```  
As expected, trajectory of site '2' is less straight than trajectory of site '1'. The function does not perform the Mantel test, because temporal autocorrelation of trajectory positions biases the test.

When trajectories have been sampled the same number of times, function `trajectoryConvergence` allows performing tests of convergence based on the trend analysis of the sequences of distances between points of the two trajectories (i.e. first-first, second-second, ...):

```{r} 
trajectoryConvergence(D, sites, surveys, symmetric = TRUE)
```  
Actually, the function performs the Mann-Whitney trend test. Values of the statistic ('tau') larger than 0 correspond to trajectories that are diverging, whereas values lower than 0 correspond to trajectories that are converging. By setting `symmetric = FALSE` the convergence test becomes asymmetric. In this case the sequence of distances is that of points of one trajectory projected onto the other:

```{r} 
trajectoryConvergence(D, sites, surveys, symmetric = FALSE)
```  


### 2.6 Distances between segments and trajectories

To start comparing trajectories between sites, one important step is the calculation of distances between directed segments, which can be obtained by calling function `segmentDistances`:
```{r} 
segmentDistances(D, sites, surveys)$Dseg
```  
Distances between segments are affected by differences in both *position* and *direction*. Hence, among the six segments of this example, the distance is maximum between the first segment of trajectory '1' and the first segment of trajectory '2'.

Distances between segments are internally calculated when comparing whole trajectories using function `trajectoryDistances`. Here we show the dissimilarity between the two trajectories as assessed using either the *Hausdorff* distance (equal to the maximum distance between directed segments) or the *directed segment path distance* (an average of distances between segments): 
```{r} 
trajectoryDistances(D, sites, surveys, distance.type = "Hausdorff")
trajectoryDistances(D, sites, surveys, distance.type = "DSPD")
```  

## 3. Structural dynamics in permanent plots

In this example we analyze the dynamics of 8 permanent forest plots located on slopes of a valley in the New Zealand Alps. The study area is mountainous and centered on the Craigieburn Range (Southern Alps), South Island, New Zealand.  Forests plots are almost monospecific, being the mountain beech (*Fuscospora cliffortioides*) the main dominant tree species. Previously forests consisted of largely mature stands, but some of them were affected by different disturbances during the sampling period (1972-2009) which includes 9 surveys. We begin our example by loading the data set, which includes 72 plot observations:
```{r load avoca, echo=T}
data("avoca")
```
Community data is in form of an object `stratifiedvegdata`. To account for differences in tree diameter, while emphasizing regeneration, the data contains individual counts to represent tree abundance and trees are classified in 19 quadratic diameter bins (in cm): {(2.25, 4], (4, 6.25], (6.25, 9], ... (110.25, 121]}. The data set also includes vectors `avoca_surveys` and `avoca_sites` that indicate the survey and forest plot corresponding to each forest state.

Before starting community trajectory analysis, we have to use function `vegdiststruct`to calculate distances between forest plot states in terms of structure and composition:
```{r distance, echo=TRUE}
avoca_D_man = vegdiststruct(avoca_strat, method="manhattan", transform = function(x){log(x+1)})
```
Distances in `avoca_D_man` are calculated using the Manhattan metric. 

### 3.1 Display trajectories in PCoA

The distance matrix `avoca_D_man` conforms our definition of $\Omega$. We use `trajectoryPCoA` to display the relations between forest plot states in this space and to draw the trajectory of each plot:

```{r avoca pcoa, echo=T, fig=TRUE, fig.height=6, fig.width=6, fig.align = "center"}
trajectoryPCoA(avoca_D_man,  avoca_sites, avoca_surveys,
               traj.colors = brewer.pal(8,"Accent"), 
               axes=c(1,2), length=0.1, lwd=2)
legend("topright", bty="n", legend = 1:8, col = brewer.pal(8,"Accent"), lwd=2)

```
Note that in this case, the full $\Omega$ include more than two dimensions, and PCoA is representing 43% of total variance (correction for negative eigenvalues is included in the call to `cmdscale` from `trajectoryPCoA`), so one has to be careful when interpreting trajectories visually.

```{r int1, echo=FALSE}
plotTrajDiamDist<-function(cli = 7) {
l = colnames(avoca_strat[[1]])
ncl = 14
m197072= avoca_strat[avoca_surveys==1][[cli]]["NOTCLI",2:ncl]
m197072[m197072<1] = NA
m1974 = avoca_strat[avoca_surveys==2][[cli]]["NOTCLI",2:ncl]
m1974[m1974<1] = NA
m1978 = avoca_strat[avoca_surveys==3][[cli]]["NOTCLI",2:ncl]
m1978[m1978<1] = NA
m1983 = avoca_strat[avoca_surveys==4][[cli]]["NOTCLI",2:ncl]
m1983[m1983<1] = NA
m1987 = avoca_strat[avoca_surveys==5][[cli]]["NOTCLI",2:ncl]
m1987[m1987<1] = NA
m1993 = avoca_strat[avoca_surveys==6][[cli]]["NOTCLI",2:ncl]
m1993[m1993<1] = NA
m1999 = avoca_strat[avoca_surveys==7][[cli]]["NOTCLI",2:ncl]
m1999[m1999<1] = NA
m2004 = avoca_strat[avoca_surveys==8][[cli]]["NOTCLI",2:ncl]
m2004[m2004<1] = NA
m2009 = avoca_strat[avoca_surveys==9][[cli]]["NOTCLI",2:ncl]
m2009[m2009<1] = NA


plot(m197072, type="l", ylim=c(1,200), log="y",
       xlab="", ylab="Number of individuals (log)", main=paste0("Trajectory ",cli), 
       axes=FALSE, col=gray(0.8), lwd=2)
axis(2, las=2)
axis(1, at=1:(ncl-1), labels=l[2:ncl], las=2)
lines(m1974, col=gray(0.7), lwd=2)
lines(m1978, col=gray(0.6), lwd=2)
lines(m1983, col=gray(0.5), lwd=2)
lines(m1987, col=gray(0.4), lwd=2)
lines(m1993, col=gray(0.3), lwd=2)
lines(m1999, col=gray(0.2), lwd=2)
lines(m2004, col=gray(0.1), lwd=2)
lines(m2009, col=gray(0), lwd=2)
legend("topright", bty="n", lwd=2,col=gray(seq(0.8,0, by=-0.1)), legend=c("1970/72","1974","1978","1983", "1987", "1993","1999","2004","2009"))
}
```

One can inspect specific trajectories using argument `selection` in function `trajectoryPCoA`. This allows getting a better view of particular trajectories, here that of forest plot '3':
```{r trajectory 3 DBH dist, fig.height=4, fig.width=8, fig.align = "center"}
par(mfrow=c(1,2))
trajectoryPCoA(avoca_D_man,  avoca_sites, avoca_surveys,
               selection= 3,
               length=0.1, lwd=2)
plotTrajDiamDist(3)
```

In the right hand, we added a representation of the change in the mountain beech tree diameter distribution through time for trajectory of forest plot '3'. The dynamics of this plot include mostly growth, which results in individuals moving from one diameter class to the other. The whole trajectory looks mostly directional. Let's now inspect the trajectory of forest plot '4':
```{r trajectory 4, echo=T, fig.height=4, fig.width=8, fig.align = "center"}
par(mfrow=c(1,2))
trajectoryPCoA(avoca_D_man,  avoca_sites, avoca_surveys,
               selection= 4,
               length=0.1, lwd=2)
plotTrajDiamDist(4)
```

This second trajectory is less straight and seems to include a turn by the end of the sampling period, corresponding to the recruitment of new saplings.

### 3.2 Trajectory lengths, angles and directionality

While trajectory lengths and angles can be inspected visually in ordination diagrams, it is better to calculate them using the full $\Omega$ space (i.e., from matrix `avoca_D_man`). Using function `trajectoryLengths` we can see that the trajectory of forest plot '4' is lengthier than that of plot '3', mostly because includes a lengthier last segment (i.e. the recruitment of new individuals):

```{r trajectory lengths, echo=T}
trajectoryLengths(avoca_D_man, avoca_sites, avoca_surveys)
```

If we calculate the angles between consecutive segments (using function `trajectoryLengths`) we see that indeed the trajectory of '3' is rather directional, but the angles of trajectory of '4' are not that low:

```{r trajectory angles, echo=T}
trajectoryAngles(avoca_D_man, avoca_sites, avoca_surveys)
```

By calling function `trajectoryDirectionality` we can confirm that the trajectory for site '4' is less straight than that of site '3':

```{r trajectory directionality, echo=T}
trajectoryDirectionality(avoca_D_man, avoca_sites, avoca_surveys)
```


### 3.3 Distances between trajectories

We can calculate the resemblance between forest plot trajectories using `trajectoryDistances`:
```{r avoca DT, echo=FALSE}
avoca_D_traj_man = trajectoryDistances(avoca_D_man, avoca_sites, distance.type="DSPD", verbose=FALSE)
print(round(avoca_D_traj_man,3))
```
The closest trajectories are those of plots '1' and '2'. They looked rather close in *position* in the PCoA ordination of $\Omega$ with all trajectories, so probably it is position, rather than *shape* which has influenced this low value. The next pair of similar trajectories are those of the '3'-'5' pair. We can use `cmdscale` to produce an ordination of resemblances between trajectories:
```{r avoca DT PCoA, echo=TRUE, fig = TRUE, fig.height=5, fig.width=6, fig.align="center"}
cmd_D2<-cmdscale(avoca_D_traj_man, add=TRUE, eig=TRUE, k=7)
x<-cmd_D2$points[,1]
y<-cmd_D2$points[,2]
plot(x,y, type="p", asp=1, xlab=paste0("PCoA 1 (", round(100*cmd_D2$eig[1]/sum(cmd_D2$eig)),"%)"), 
     ylab=paste0("PCoA 2 (", round(100*cmd_D2$eig[2]/sum(cmd_D2$eig)),"%)"), col="black",
     bg= brewer.pal(8,"Accent"), pch=21)
text(x,y, labels=1:8, pos=2)
```

